\texorpdfstring {Implementing Equational Constraints \\ in a Functional Language} {Implementing Equational Constraints in a Functional Language}

### 2012-11-08 papers/wlp2011/paper.tex
Modifications to the PADL paper

In the next section, we review the source language Curry
and the features considered in this paper.
Section~\ref{sec:Compilation} sketches the implementation scheme
of KiCS2.
Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
discuss the extensions to implement unification modulo
functional evaluation and functional patterns, respectively.
Benchmarks demonstrating the usefulness of this scheme
are presented in Sect.~\ref{sec:Benchmarks}
before we conclude in Sect.~\ref{sec:Conclusions}.

@@

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
where \code{Fail} represents a failure and \code{(Choice i t1 t2)}
a non-deterministic value, i.e.,
a selection of two values \code{t1} and \code{t2}
that can be chosen by some search strategy.
The first argument \code{i} of type \code{ID} of a \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Sect.~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice} constructor. The difficulty is to get a unique identifier
on demand, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick its unique identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple one) and our system
is parametric over concrete implementations of \code{IDSupply}.

@@

Since all data types are extended by additional constructors,
we must also extend the definition of operations performing
pattern matching.\footnote{To obtain a simple compilation scheme,
KiCS2 transforms source programs into uniform programs
\cite{BrasselHanusPeemoellerReck11} where pattern matching
is restricted to a single argument.
This is always possible by introducing auxiliary operations.}
For instance, consider the definition of polymorphic lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is then extended as described above:
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
%
The second rule transforms a non-deterministic argument
into a non-deterministic result and
the final rule returns \code{Fail} in all other cases,
i.e., if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation (failure propagation).

@@

The requirement to make consistent decisions can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice has
not been decided yet. Furthermore, we assume operations to lookup the 
current decision for a given identifier or change it
(depending on the implementation
of \code{IDSupply}, KiCS2 supports several implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()$\listline$
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
    where
      follow ChooseLeft  = printValsDFS x1
      follow ChooseRight = printValsDFS x2
      follow NoDecision  = do newDecision ChooseLeft  x1
                              newDecision ChooseRight x2$\listline$
      newDecision d x = do setDecision i d
                           printValsDFS x
                           setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{haskell}
This operation ignores failures and prints values that are not rooted by a
\code{Choice} constructor. For a \code{Choice} constructor, 
it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing all values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

Other search strategies, like
breadth-first search, iterative deepening, or parallel search,
can be obtained by different implementations of this main operation
to print all values.
Instead of printing them, one can also collect
the values in a tree-like data structure
so that the programmer can implement his own search strategies
as tree traversals. Thus, KiCS2 supports a primitive
\begin{curry}
  getSearchTree :: a -> IO (SearchTree a)
\end{curry}
that returns the search tree corresponding to the evaluation
of its argument, where the search tree is some computed value,
a failure, or a choice between two trees:
\begin{curry}
  data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
This primitive is useful to encapsulate non-deterministic
operations and select some result value, e.g., the
``first'' or the ``best'' one according to some ordering.
Since the search tree is created in a demand-driven manner,
the primitive is also applicable to infinite search spaces
(in contrast to Prolog's \code{findall} primitive \cite{Naish85SLP}).
Based on this representation, a Curry programmer
can define his own search strategies
as tree traversals in his source program without any modification
of the Curry compiler (see \cite{HanusPeemoellerReck12ATPS}
for detailed examples).
Note that these kinds of applications demand for a side-effect
free implementation of non-deterministic computations
(in contrast to traditional Prolog implementations
\cite{Ait-Kaci91})---which is the challenge addressed in this paper.

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, KiCS2 performs an optimization for
deterministic operations. If an operation is defined by non-overlapping
rules and does not call, neither directly nor indirectly through
other operations, a function defined by overlapping rules,
the evaluation of such an operation (like \code{xor} or \code{not})
cannot introduce non-deterministic values.
Thus, it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended by additional cases
for handling \code{Choice} and \code{Fail} so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
show that for deterministic operations this implementation outperforms 
all other Curry implementations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

@@

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (we only present the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success

@@

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by overlapping rules)
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, in the following, we consider the
implementation of \code{IDSupply} to be unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} identifies a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
%
As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this identifier:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
A single \code{Constraint} provides the decision for
one constructor. In order to support constraints for structured data,
a list of \code{Constraint}s provides the decision for the outermost
constructor and the decisions for all its arguments.
Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent with the decisions previously made during search.
These binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decisions as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True  = Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False = Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:%
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard [i :=: BindTo j] Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard cs x) = do consistent <- add cs
                                 if consistent then do printValsDFS x
                                                       remove cs
                                               else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraints \code{cs}
with the decisions made so far and, in case of consistency,
stores the decisions made by the constraints.
In this case, the constrained value is evaluated
before the constraints are removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

@@

Our proposed implementation of functional patterns in KiCS2
has a structure that is quite similar to that of equational constraints
with the exception that variables could be also bound
to unevaluated expressions. 
Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Sect.~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = [i :=: ChooseLeft]
  lazyBind i False = [i :=: ChooseRight]
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

>>>>

In the next section, we review the source language Curry
and the features considered in this paper.
Section~\ref{sec:Compilation} recapitulates the implementation scheme
of KiCS2 originaly presented in \cite{BrasselHanusPeemoellerReck11}.
Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
discuss our new extensions to implement unification modulo
functional evaluation and functional patterns, respectively.
Benchmarks demonstrating the usefulness of these extensions
are presented in Sect.~\ref{sec:Benchmarks}
before we conclude in Sect.~\ref{sec:Conclusions}.

@@

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
where \code{Fail} represents a failure and \code{(Choice i t1 t2)}
a non-deterministic value, i.e.,
a selection of two values \code{t1} and \code{t2}
that can be chosen by some search strategy.
The first argument \code{i} of type \code{ID} of a \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Sect.~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice} constructor. The difficulty is to get a unique identifier
on demand, i.e., when some operation evaluates to a \code{Choice}.
We cannot thread an identifier supply, e.g. a counter,
through the search tree without fixing an evaluation order.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we can neither use unsafe features with side effects to generate
such identifiers. 
Hence, we follow the idea presented in \cite{AugustssonRittriSynek94} and
pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick its unique identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available
(see below for a simple one) and our system
is parametric over concrete implementations of \code{IDSupply}.

@@

Since all data types are extended by additional constructors,
we must also extend the definition of operations performing
pattern matching.\footnote{To obtain a simple compilation scheme,
KiCS2 transforms source programs into uniform programs
\cite{BrasselHanusPeemoellerReck11} where pattern matching
is restricted to a single argument.
This is always possible by introducing auxiliary operations.}
For instance, consider the definition of polymorphic lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is then extended as described above:
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
%
The second rule transforms a non-deterministic argument
into a non-deterministic result and
the final rule returns \code{Fail} in all other cases,
i.e., if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation (failure propagation).
Since \code{head} is a deterministic operation, it does
not use the identifier supply \code{s}.

@@

The requirement to make consistent decisions can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice has
not been decided yet. Furthermore, we assume operations to lookup the 
current decision for a given identifier or change it
(depending on the implementation
of \code{IDSupply}, KiCS2 supports several implementations
based on memory cells or finite maps).
For a top level operation that prints all values contained in a choice structure
in a depth-first manner, these operations would be of the following types:
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now the search operation can be defined by the I/O operation below:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()$\listline$
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
    where
      follow ChooseLeft  = printValsDFS x1
      follow ChooseRight = printValsDFS x2
      follow NoDecision  = do newDecision ChooseLeft  x1
                              newDecision ChooseRight x2$\listline$
      newDecision d x = do setDecision i d
                           printValsDFS x
                           setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{haskell}
This operation ignores failures and prints values that are not rooted by a
\code{Choice} constructor. For a \code{Choice} constructor, 
it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing all values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

Other search strategies, like
breadth-first search, iterative deepening, or parallel search,
can be obtained by different implementations of this main operation
to print all values.
Instead of printing them, one can also collect
the values in a tree-like data structure
for further processing. Thus, KiCS2 supports a primitive
\begin{curry}
  getSearchTree :: a -> IO (SearchTree a)
\end{curry}
that returns the search tree corresponding to the evaluation
of its argument, where the search tree is some computed value,
a failure, or a choice between two trees:
\begin{curry}
  data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
This primitive is useful to encapsulate non-deterministic
operations and select some result value, e.g., the
``first'' or the ``best'' one according to some ordering.
Since the search tree is created in a demand-driven manner,
the primitive is also applicable to infinite search spaces
(in contrast to Prolog's \code{findall} primitive \cite{Naish85SLP}).
Based on this representation, a Curry programmer
can define his own search strategies
as tree traversals in his source program without any modification
of the Curry compiler (see \cite{HanusPeemoellerReck12ATPS}
for detailed examples).
Note that these kinds of applications demand for a side-effect
free implementation of non-deterministic computations
(in contrast to traditional Prolog implementations
\cite{Ait-Kaci91})---which is the challenge addressed in this paper.

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, KiCS2 performs an optimization for
deterministic operations. If an operation is defined by non-overlapping
rules and does not call, neither directly nor indirectly through
other operations, a function defined by overlapping rules,
the evaluation of such an operation (like \code{head})
cannot introduce non-deterministic values.
Thus, it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended by additional cases
for handling \code{Choice} and \code{Fail} so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
show that for deterministic operations this implementation outperforms 
all other Curry implementations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

@@

Due to this definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (we only present the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success

@@

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by overlapping rules)
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, in the following, we consider the
implementation of \code{IDSupply} to be unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} identifies a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
We use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
to ensure that the resulting choice is not considered a free variable.

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this identifier:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
A single \code{Constraint} provides the decision for
one constructor. In order to support constraints for structured data,
a list of \code{Constraint}s provides the decision for the outermost
constructor and the decisions for all its arguments.
Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent with the decisions previously made during search.
These binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decisions as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True  = Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False = Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:%
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard [i :=: BindTo j] Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard cs x) = do consistent <- add cs
                                 if consistent then do printValsDFS x
                                                       remove cs
                                               else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraints \code{cs}
with the decisions made so far and, in case of consistency,
stores the decisions made by the constraints.
In this case, the constrained value is evaluated
before the constraints are removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

@@

Our proposed implementation of functional patterns in KiCS2
has a structure that is quite similar to that of equational constraints
with the exception that variables could be also bound
to unevaluated expressions. 
Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Sect.~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = [i :=: ChooseLeft]
  lazyBind i False = [i :=: ChooseRight]
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} 
as far as necessary when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.
